---
title: Analyisis to determine economic impact of severe weather events using NOAA
  storm data
author: "Telvis Calhoun"
date: "December 26, 2015"
output:
  html_document:
    keep_md: true
---

## Introduction

Title: Analyisis to determine economic impact of severe weather events using NOAA storm data

Synopsis: In this document, we analyze the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database to determine the economic impact of different weather events. The goal of the analysis is to rank the storm events based on their economic impact. Our hope is that this ranking will help policy-makers to allocate emergency management funds to the events with the greatest economic impact.  

## Data Processing


First, lets load libraries used in the analysis

```{r, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```

Let's load the data
```{r,echo=TRUE}
df <- read.csv("activity.csv", na.strings = "NA", stringsAsFactors = FALSE)
```

TODO...


## Results
TODO....
- At most 3 plots


